---
title: "p8105_hw3_ar4459"
author: "ASHLEY ROMO"
date: "2023-10-13"
output: github_document
---

```{r}
library(tidyverse)
library(patchwork)

# setting the width and height of the plots
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```


## Problem 1

```{r}

library(p8105.datasets)

data("instacart")
```

There are a total of 1,384,617 rows and 15 columns. The structure of the data is contains many ids including order_id, product_id, user_id, aisle_id, and department_id. Some key variables include product names, order_hour_of_day, and reordered because they describe at what time during the day specific products are purchased or reordered, which can help the store advertise the most sold items during the hours they are most often purchased or reordered.

```{r}
instacart_df = instacart
 
# number of aisles 
total_aisle = instacart_df |> 
  select(aisle) |> distinct()
nrow(total_aisle)


# aisles with the most items ordered from
most_ordered = instacart_df |> 
  group_by(aisle) |> 
  summarise(n_count=n()) |> 
  arrange(-n_count) |> 
  filter(row_number() == 1)
```


The total number of aisles is `r nrow(total_aisle)`. The aisle with the most ordered from is `r most_ordered`.

### Problem 1 (ANSWER KEY)

#### Read in the data

```{r}
data("instacart")

instacart = 
  instacart |> 
  as_tibble()
```

#### Answer questions about the data

This dataset contains `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row resprenting a single product from an instacart order. Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. In total, there are `r instacart |> select(product_id) |> distinct() |> count()` products found in `r instacart |> select(user_id, order_id) |> distinct() |> count()` orders from `r instacart |> select(user_id) |> distinct() |> count()` distinct users.

Below is a table summarizing the number of items ordered from aisle. In total, there are 134 aisles, with fresh vegetables and fresh fruits holding the most items ordered by far.

```{r}
instacart |> 
  count(aisle) |> 
  arrange(desc(n))
```

Next is a plot that shows the number of items ordered in each aisle. Here, aisles are ordered by ascending number of items.

```{r}
instacart |> 
  count(aisle) |> 
  filter(n > 10000) |> 
  mutate(aisle = fct_reorder(aisle, n)) |> 
  ggplot(aes(x = aisle, y = n)) + 
  geom_point() + 
  labs(title = "Number of items ordered in each aisle") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```

Our next table shows the three most popular items in aisles `baking ingredients`, `dog food care`, and `packaged vegetables fruits`, and includes the number of times each item is ordered in your table.

```{r}
instacart |> 
  filter(aisle %in% c("baking ingredients", "dog food care", "packaged vegetables fruits")) |>
  group_by(aisle) |> 
  count(product_name) |> 
  mutate(rank = min_rank(desc(n))) |> 
  filter(rank < 4) |> 
  arrange(desc(n)) |>
  knitr::kable()
```

Finally is a table showing the mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week. This table has been formatted in an untidy manner for human readers. Pink Lady Apples are generally purchased slightly earlier in the day than Coffee Ice Cream, with the exception of day 5.

```{r}
instacart |>
  filter(product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>
  group_by(product_name, order_dow) |>
  summarize(mean_hour = mean(order_hour_of_day)) |>
  pivot_wider(
    names_from = order_dow, 
    values_from = mean_hour) |>
  knitr::kable(digits = 2)
```


#Problem 2

First, I load the BRFSS data.

```{r}
library(p8105.datasets)
data("brfss_smart2010")
```

Now, I will do some data cleaning

```{r}
brfss_df = 
  brfss_smart2010 |> 
  janitor::clean_names() |> 
  select(
    year, state = locationabbr, county = locationdesc, everything()
    ) |> 
  filter(
    topic == "Overall Health",
    response == "Poor" | response == "Fair"| response == "Good"| response == "Very good"| response == "Excellent"
    ) |> 
  mutate(
    response = factor(response, levels = c("Poor", "Fair", "Good", "Very good", "Excellent"))
    )

```

```{r}
#states observed at 7 or more locations in 2002
states_2002 =
  brfss_df |> 
  filter(year == 2002) |> 
  select(year, state, county) |> 
  distinct(year, state, county) |> 
  group_by(year, state) |> 
  summarize(n_obs = n()) |> 
  filter(n_obs >= 7) 
  
nrow(states_2002)
  

#states observed at 7 or more location in 2010
states_2010 =
  brfss_df |> 
  filter(year == 2010) |> 
  select(year, state, county) |> 
  distinct(year, state, county) |> 
  group_by(year, state) |> 
  summarize(n_obs = n()) |> 
  filter(n_obs >= 7) 

nrow(states_2010)
```

In 2002, `r nrow(states_2002)` states were observed at 7 or more locations. These states include CT, FL, MA, NC, NJ, and PA. 

In 2010, `r nrow(states_2010)` states were observed at 7 or more locations. These states include CA, CO, FL, MA, MD, NC, NE, NJ, NY, OH, PA, SC, TX, and WA. 


Spaghetti plot of the average  data value over time within a state.

```{r}
excellent_df = 
  brfss_df |> 
  filter(response == "Excellent") |> 
  select(year, state, data_value) |> 
  group_by(year, state) |> 
  summarise(
   avg_data_val = mean(data_value)
  ) 

excellent_plot = 
  excellent_df |> 
  ggplot(aes(x = year, y = avg_data_val, color = state)) +
  geom_line() + 
  theme(legend.position ="right") +
  labs(
    title = "Average data value over time",
    x = "Year",
    y = "Average data value",
    color = "State"
  )


excellent_plot

```

The plot shows the average value of data over time for each state. We see there is an overall slight negative trends. As time passes by, the average data value decreases slightly. However, the plot also shows the that each state has variable average data value throughout the year with sharp increases and deceases. 



Two-panel plot

```{r}
val_06_10_df =
  brfss_df |> 
  filter(year == 2006 | year == 2010, state == "NY") |> 
  group_by(data_value, response, county) 

val_06_10_plot = 
  val_06_10_df |> 
  ggplot(aes(x = response, y = data_value)) +
  geom_boxplot() +
  labs(
    title = "2006 vs. 2010",
    color = "Year",
    x = "Response",
    y = "Value"
  ) +
  facet_grid(. ~ year)

val_06_10_plot

```
The two panel plots shows us similarities between 2006 and 2010. For example, we see that those that responded  that their health was poor had a low data value. Those that responded they had excellent health had lower data value than those that responded good or very good; this was see in 2006 and 2010. Overall, the data value aligned with the response provided regarding their health except for those who responded they had excellent health. 


## Problem 3

First, load the data

```{r}
accel_df = read_csv("data/nhanes_accel.csv") |> 
  janitor::clean_names() |> 
  pivot_longer(
    min1:min1440, 
    names_prefix = "min",
    names_to = "min",
    values_to = "phys_act"
  )


covar_df = read_csv("data/nhanes_covar.csv", skip = 4) |> 
  janitor::clean_names() |> 
  drop_na() |> 
  mutate(
     sex = case_match(
       sex,
        1 ~ "male",
       2 ~ "female"
     ),
     education = case_match(
        education,
       1 ~ "Less than high school",
       2 ~ "High school equivalent",
       3 ~ "More than high school"
     )
  ) |> 
  mutate(
    education = factor(
      education, 
      levels = c("Less than high school", "High school equivalent", "More than high school")
      )
    ) |> 
  filter(age >= "21") 

```

Now, we merge the two datasets. 

```{r}
merged_df = left_join(covar_df, accel_df)
```

Tables for men and women

```{r}

#table
sex_df = 
  covar_df |>
  group_by(sex, education) |> 
  count(sex, education) |> 
  pivot_wider(
    names_from = education,
    values_from = n
  ) |> 
  knitr::kable()

sex_df

```
The table shows the number of men and women in each education category. Overall, the education category with those with more than high school education had the greatest number of individuals while the education category with those with less than high school education had the fewest number of individuals. There are more women than men in the education category of more than high school education. There are more male in the education category of those with a high school equivalent category. There are more women in the education category with less than high school education.

Visualization of the age distribution for men and women in each education category

```{r}
sex_ed_df = 
  covar_df |> 
  group_by(sex, education) 
  
sex_ed_plot =
  sex_ed_df |> 
  ggplot(aes(x = age, fill = sex)) +
  geom_density(alpha = 0.7) +
  labs(
    title = "Age Distribution For Men and Women Per Education Category",
    x = "Age",
    color = "Sex") +
  facet_grid(. ~ education) 

sex_ed_plot

```

Those with more than a high school education appear to be younger than those with a high school equivalent or less than a high school education. There are also more females than males in the more than high school education. The average age of women with more than a high school education is `r mean(pull(sex_ed_df |> filter(sex == "female", education == "More than high school"), age))` while the mean age of men with more than a high school education is `r mean(pull(sex_ed_df |> filter(sex == "male", education == "More than high school"), age))`. The mean age of those with more than a high school education is `r mean(pull(sex_ed_df |> filter(education == "More than high school"), age))`.

The average age of those with an education equivalent to a high school education is `r mean(pull(sex_ed_df |> filter(education == "High school equivalent"), age))`. The average age of women with an education equivalent to a high school degrees is `r mean(pull(sex_ed_df |> filter(sex == "female", education == "High school equivalent"), age))` while the average age of men with an education equivalent to a high school degree is `r mean(pull(sex_ed_df |> filter(sex == "male", education == "High school equivalent"), age))`.

The average age of those with less than a high school education is `r mean(pull(sex_ed_df |> filter(education == "Less than high school"), age))`. The average age of women with less than a high school education is `r mean(pull(sex_ed_df |> filter(sex == "female", education == "Less than high school"), age))` while the average age of men with less than a high school education is `r mean(pull(sex_ed_df |> filter(sex == "male", education == "Less than high school"), age))`.

Total activity for each participant

```{r}
total_activity_df =
  merged_df |> 
  group_by(seqn, age, sex, education) |> 
  summarize(
    total_act = sum(phys_act)) |>
  drop_na() 

total_activity_plot = 
  total_activity_df |> 
  ggplot(aes(x = age, y = total_act,  color = sex)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  labs(
    title = "Total Activity For Each Participant Over The Day",
    x = "Age",
    y = "Total Activity",
    color = "Sex"
  ) +
  facet_grid(. ~ education)

total_activity_plot
```
Across all education levels, total activity decreases as age increases. Among those with a high school equivalent or more than a high school education, women appear to have a higher total activity compared to men. Among those with less than a high school education, men appear to have higher total activity. 

Those with less than a high school education have a average age of `r mean(pull(total_activity_df |> filter(education == "Less than high school"), age))`. Within this education category, women have an average age of `r mean(pull(total_activity_df |> filter(sex == "female", education == "Less than high school"), age))` and men have an average age of `r mean(pull(total_activity_df |> filter(sex == "female", education == "Less than high school"), age))`.

Those with a high school equivalent education have an average age of `r mean(pull(total_activity_df |> filter(education == "High school equivalent"), age))`. Within this education cateogry, women have an average age of `r mean(pull(total_activity_df |> filter(sex == "female", education == "High school equivalent"), age))` and men have an average age of `r mean(pull(total_activity_df |> filter(sex == "male", education == "High school equivalent"), age))`.

Lastly, those with more than a high school education have an average age of `r mean(pull(total_activity_df |> filter(education == "More than high school"), age))`. Within this category, women have an average age of `r mean(pull(total_activity_df |> filter(sex == "female", education == "More than high school"), age))` and men have an average age of `r mean(pull(total_activity_df |> filter(sex == "male", education == "More than high school"), age))`.


24-hour activity time courses for each education level by sex

```{r}
merged_df |> 
  drop_na() |> 
  ggplot(aes(x = min, y = phys_act,  color = sex, group = seqn)) +
  geom_line( alpha = 0.2) +
  facet_grid(. ~ education) +
  geom_smooth(aes(group = sex), se = FALSE) +
  labs(
    title = "24-hour Activity For Each Education Level By Sex",
    x = "Minutes",
    y = "Activity"
  )
```

The trend in activity is similar across all education levels. The activity increases, makes a share decrease, and then increase again across all education levels. I conclude that individuals across education level have a similar activity level. 

The mean activity for those with less than high school education is `r mean(pull(merged_df |> drop_na() |> filter(education == "Less than high school"), phys_act))`. The mean for those with a high school equivalent education is `r mean(pull(merged_df |> drop_na() |> filter(education == "High school equivalent"), phys_act))`. The activity for those with more than a high school education is `r mean(pull(merged_df |> drop_na() |> filter(education == "More than high school"), phys_act))`.
